\documentclass{assignment}
\begin{document}

\title{5370 Midterm}


\textbf{5.} \label{NewToyYoda}
Continuation from Problem~\ref{SquishyBear}.
The distribution of English word length, $k$, is a shifted Poisson random variable. At the end of each word, we will add
a space to separate words. There are no spaces allowed in the middle of a word. Under this assumption word length
follows the distribution
\begin{align}
  w(k)=\frac{e^{-\lambda}\lambda^{k-2} }{(k-2)!} ; k=2,3,4,\cdots \label{eqn:problem5:length}
\end{align}
$$ $$
Assuming there are about a million words in the English language.
\begin{enumerate}
\item What is the probability of generating a word in the English language randomly choosing letters
  \begin{itemize}
  \item assuming the letters are chosen uniformly.
  \item assuming the letters are chosen according to $p(x)$.\footnote{See Problem~\ref{SquishyBear}.}
  \end{itemize}
  \textbf{Solution:} \\
  \begin{align*}
    \text{Let} \quad & \begin{cases}
      K & = \text{probability of generating a sequence of non-space alphabets followed by a space
      character} \\
      E & = \text{probability of generating a valid English word} \\
      N & = \text{total number of English words}, 1000000
    \end{cases}
  \end{align*}
  Using conditional probability relation we can express $\Pr\{E\}$ as:
  $$\Pr\{E\}= \sum_k \Pr\{E \cap K\} = \sum_k \Pr\{E|K=k\}\Pr\{K=k\}$$
  where
  $$\Pr\{K=k\} = (1 - \Pr\{\text{space}\})^{k-1}\Pr\{\text{space}\}$$

  Since $w(k) * N \approx $ number of English words with length $k$, we can express $\Pr\{E|K=k\}$ as
  $$\Pr\{E|K=k\} \approx \df{w(k) * N}{26^{k-1}\times 1}$$

  However, when $k=2,3,4$, using (\ref{eqn:problem5:length}) gives us unexpected word counts:
  \begin{align*}
    w(2) & = \df{e^{-6.1}6.1^0}{0!} \approx 0.0060967 \\
    w(2) * 1000000 &  \approx 2242 > 26 \\
    w(3) & = \df{e^{-6.1}6.1^1}{1!} \approx 0.013681 \\
    w(3) * 1000000 & \approx 13681 > 26^2 \\
    w(4) & = \df{e^{-6.1}6.1^2}{2!} \approx 0.041729 \\
    w(4) * 1000000 & \approx 41728 > 26^3
  \end{align*}
  Note that I am using $\lambda = 6.1$ since we are adding an additional space at the end of each word.


  For this reason, in order to get a reasonable estimation, I will use some data I found online for $k=2,3,4$ terms.

  According to Wikipedia,
  \begin{itemize}
  \item Number of 1 letter-words = 3\\
    source: \url{https://en.wiktionary.org/wiki/Category:English_one-letter_words}
  \item Number of 2 letter-words = 114\\
    source: \url{https://en.wiktionary.org/wiki/Category:English_two-letter_words}
  \item Number of 3 letter-words = 172\\
    source: \url{https://en.wiktionary.org/wiki/Category:English_three-letter_words}
  % \item Number of 4 letter not-words = 2745\\
  %   source: \url{http://www.appsapps.info/stuff/notwords4.txt}
  \end{itemize}

  Thus, my estimation for generating a word in English is:
  \begin{align*}
    \Pr\{E\} & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}}
               (1 - \Pr\{\text{space}\})^{k-1}\Pr\{\text{space}\}\right) \\
    \Pr\{E\} & = \df{3}{26}(1 - \Pr\{\text{space}\})\Pr\{\text{space}\} + \\
             & \quad \df{114}{26^2}(1 - \Pr\{\text{space}\})^2\Pr\{\text{space}\} + \\
             & \quad \df{172}{26^3}(1 - \Pr\{\text{space}\})^3\Pr\{\text{space}\} + \\
             & \quad \sum_{k=5}^{N} \left(\df{w(k) * N}{26^{k-1}}
               (1 - \Pr\{\text{space}\})^{k-1}\Pr\{\text{space}\}\right)
  \end{align*}
  % For $\Pr\{E|K=k\}$, it's the probability of English word when given a sequence of non-space alphabets
  % followed by a space. So it can be computed with number of English words in a given length divided by
  % all the possible combination of non-space alphabets in that given length. Since we know the total
  % number of English words and the
  So,
  \begin{itemize}
    \item If letters are chosen uniformly,
      \begin{align*}
        \Pr\{\text{space}\}
        & = \df{1}{27} \\
        \Pr\{E\}
        & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}}\left( \df{26}{27} \right)^{k-1}\df{1}{27}\right) \\
        & = \sum_{k=2}^{N} \left(\df{w(k) * N}{27^{k}} \right) \\
        & = \df{3}{27^2} + \df{114}{27^3} + \df{172}{27^4} +
          \sum_{k=5}^{N} \left(\df{w(k) * N}{27^{k}} \right) \\
        & \approx 0.01649 \numberthis \label{result:prob5:1}
      \end{align*}
    \item If letters are chosen according to $p(x)$,
      \begin{align*}
        \Pr\{\text{space}\}
        & = \df{40000}{222303} \approx 0.17993 \\
        \Pr\{E\} & = \df{3}{26} \df{182303}{222303} \df{40000}{222303} +
                   \df{114}{26^2} \left(\df{182303}{222303}\right)^2 \df{40000}{222303} +
                   \df{172}{26^3} \left(\df{182303}{222303}\right)^3 \df{40000}{222303} + \\
        & \quad \sum_{x=5}^{N} \left( \df{w(k) * N}{26^{k-1}}
          \left(\df{182303}{222303}\right)^{k-1} \df{40000}{222303} \right) \\
        & \approx 0.054269 \numberthis \label{result:prob5:2}
      \end{align*}

      \CodeSnippet{clojure}
      {Code used to compute the probability in (\ref{result:prob5:1}) and (\ref{result:prob5:2})}
      {lst:prob5:1}
      {../../src/elc5370/problem5.clj}{44}{68}

      % \begin{align*}
      %   P(W) & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}} (1 - p(0))^{k-1}p(0)\right) \\
      %   P(W) & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}}
      %          \left(1 - (1-\alpha)\alpha^0\right)^{k-1}(1-\alpha)\alpha^0\right) \\
      %        & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}} \alpha^{k-1}(1-\alpha)\right) \\
      %        & = \sum_{k=2}^{N} \left(\df{w(k) * N}{26^{k-1}} p(k-1)\right)
      % \end{align*}
  \end{itemize}
  % % \url{http://www.languagemonitor.com/number-of-words/number-of-words-in-the-english-language-1008879/}
  % According to \url{http://www.languagemonitor.com}, $N \approx 1,025,109$

  % 4 letters non words \url{http://www.appsapps.info/stuff/notwords4.txt}

\item What is the probability of randomly generating $W$ words in a row under each distribution?\\
  \textbf{Solution:}\\
  \begin{itemize}
  \item If letters are chosen uniformly: $0.01649^W$
  \item If letters are chosen according to $p(x)$: $0.054269^W$
  \end{itemize}

\item How many times must we randomly sample before getting $W$ words in a row under each distribution? \\
  \textbf{Solution:} \\
  Given a event's probability of occurrence is $p$, the expected number of trial to the first success should be
  $\df{1}{p}$. Thus
  \begin{itemize}
    \item Uniform distribution: we need $\ceil{0.01649^{-W}}$ number of trials.
    \item For $p(x)$, we need $\ceil{0.054269^{-W}}$ number of trials.
  \end{itemize}
  % According to AEP,
  % \begin{align*}
  %   H(E) & \approx - \df{1}{W}\log p(E_1, E_2, E_3, \cdots, E_W) \\
  %        & = - \df{1}{W} \log 0.01649^W = - \log 0.01649\\
  % \end{align*}
  % We probably need to sample $2^{WH}$ times which gives us $2^{-W\log 0.01649} = \left(\df{1}{0.01649}\right)^W$ which
  % is for uniform distribution. For distribution using $p(x)$, we would need $\ceil{\left(\df{1}{0.054269}\right)^W}$

\item If a word is $w_k$ characters long (including a space at the end), assume it requires $k \log_2 N$ bits to form.
  How many bits do we use, on average, to query to the point we have $W$ words in a row? \\
  \textbf{Solution:} \\
  Number of bits require to encode a sequence is $W \cdot H(E)$

  Average length of random word can be calculated by:
  \begin{align*}
    \sum_{k=2}^{\infty} k(1-\Pr\{\text{space}\})^{k-1}\Pr\{\text{space}\}
  \end{align*}
  For both cases, they are computed numerically using code shown in (\ref{lst:prob5:4})
  \CodeSnippet{clojure}
  {Code used to compute expected random word length}
  {lst:prob5:4}
  {../../src/elc5370/problem5.clj}{75}{82}

  Computed results are:
  \begin{itemize}
    \item Expected random word length using uniform distribution is $26.8433 \approx 27$.
    \item Expected random word length using $p(x)$ is $5.378 \approx 6$.
  \end{itemize}

  So for uniform distribution we need $W(27\log_2N)\ceil{0.01649^{-W}}$ bits, and for $p(x)$ we would need

  $W(6\log_2N)\ceil{0.054269^{-W}}$ bits.
\end{enumerate}

\end{document}
